### 2018-07-06 paper/3-abstractions.tex
Minor revision

\section{Build systems, abstractly}\label{sec-abstractions}

\begin{figure}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store containing a key/value map and persistent build information
data Store i k v -- i = info, k = key, v = value
initialise :: i -> (k -> v) -> Store i k v
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
data Hash v -- a compact summary of a value with a fast equality check
hash    :: Hashable v => v -> Hash v
getHash :: Hashable v => k -> Store i k v -> Hash v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build tasks (see §3.2)
newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system (see §3.3)
type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system components: a scheduler and a rebuilder (see §5)
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
\vspace{-3mm}
\caption{Type signatures of key build systems abstractions.}\label{fig-types}
\vspace{-3mm}
\end{figure}

@@

\vspace{-1.5mm}
\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\vspace{-0.5mm}
% We begin by establishing a common vocabulary for build systems:

@@

\noindent
Here \hs{c} stands for \emph{constraint}, such as \hs{Applicative} or \hs{Monad}.
A \hs{Task} describes a single build task, while \hs{Tasks} associates a
\hs{Task} to every non-input key; input keys are associated with \hs{Nothing}.
The highly-abstracted type \hs{Task} describes how to build a value when given
a way to build its dependencies, and is best explained by an example.
Consider this \Excel spreadsheet:
% \footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics} might
% recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}

@@

\noindent
Here cell \cmd{A1} contains the value \cmd{10}, cell \cmd{B1} contains the
formula \cmd{A1+A2}, etc.

@@

\noindent
We can represent the formulae of this spreadsheet with the following task
description:

@@

\begin{minted}[xleftmargin=5pt]{haskell}
compute :: Task Monad k v -> Store i k v -> v
compute task store = runIdentity $ run task (\k -> return (getValue k store))
\end{minted}

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we unwrap using the function \hs{runIdentity} of type \hs{Identity}~\hs{a}~\hs{->}~\hs{a},
which both forces the monad to \hs{Identity} and strips it away.

>>>>

\section{Build systems, abstractly}\label{sec-abstractions}
\vspace{-0.5mm}

\begin{figure}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store containing a key/value map and persistent build information
data Store i k v -- i = info, k = key, v = value
initialise :: i -> (k -> v) -> Store i k v
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
\end{minted}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
data Hash v -- a compact summary of a value with a fast equality check
hash    :: Hashable v => v -> Hash v
getHash :: Hashable v => k -> Store i k v -> Hash v
\end{minted}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build tasks (see §3.2)
newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system (see §3.3)
type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system components: a scheduler and a rebuilder (see §5)
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
\vspace{-3.5mm}
\caption{Type signatures of key build systems abstractions.}\label{fig-types}
\vspace{-4.5mm}
\end{figure}

@@

\vspace{-1.5mm}
\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\vspace{-0.5mm}

@@

\noindent
Here \hs{c} stands for \emph{constraint}, such as \hs{Applicative}
(\S\ref{sec-why-polymorphism} explains why we need it).
A \hs{Task} describes a single build task, while \hs{Tasks} associates a
\hs{Task} to every non-input key; input keys are associated with \hs{Nothing}.
The highly-abstracted type \hs{Task} describes how to build a value when given
a way to build its dependencies, and is best explained by an example.
Consider this \Excel spreadsheet:
% \footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics} might
% recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}

@@

\noindent
Here cell \cmd{A1} contains the value \cmd{10}, cell \cmd{B1} contains the
formula \cmd{A1 + A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:

@@

@@

\begin{minted}[xleftmargin=5pt]{haskell}
compute :: Task Monad k v -> Store i k v -> v
compute task store = runIdentity $ run task (\k -> Identity (getValue k store))
\end{minted}

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad: the function
\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result comes as \hs{Identity}~\hs{v}, which we unwrap by \hs{runIdentity}.

% To fix the mismatch, we unwrap using the function \hs{runIdentity} of type \hs{Identity}~\hs{a}~\hs{->}~\hs{a},
% which both forces the monad to \hs{Identity} and strips it away.

### 2018-07-06 paper/4-build.tex
Minor revision

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it.
It also marks dirty all cells that depend (transitively) on the modified cell.
Excel does not record the dynamic dependencies of each cell,
so instead it computes a \emph{static over-approximation} -- it is always
safe for it to make more cells dirty than is strictly necessary, but not
vice versa.  The over-approximation is as follows: a cell is marked dirty
if its formula statically refers to a dirty cell, or if
the formula calls a function like \cmd{INDIRECT} whose dependencies cannot be
guessed from the formula alone.  The over-approximation is clear for \cmd{INDIRECT},
but it is also present for \cmd{IF}, where both branches are followed even though
dynamically only one is used.

@@

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well if we skip any
number of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$.

The deep constructive trace approach only works if the operations are
\emph{deterministic}, otherwise it is possible to violate correctness.
Concretely, if
the result of building a key given a set of immediate dependencies can vary,
then it is possible that key \hs{k} will have been computed locally, but
something \emph{using} \hs{k} will have been taken as a constructive input using
a different value for \hs{k}. If another rule combines those two values then two
different values of \hs{k} will have been used in one rule, leading to problems
such as link-errors for typical compilers -- a phenomenon known as a
\emph{Frankenbuild}~\cite{esfahani2016cloudbuild}.

A downside of deep constructive traces is that they cannot support cutoff
(\S\ref{sec-background-shake}), other than at the $n$ levels of dependencies. On
the other hand, these traces are particularly useful for ``shallow builds'', as
discussed in~\S\ref{sec-cloud-aspects}.

@@

\noindent
Here \hs{r} is either \hs{Hash}~\hs{v} (verifying traces) or
\hs{v} (constructive traces). A real system is highly likely to use a more
optimised implementation. Some of the most obvious optimisations are:

\begin{enumerate}
\item Any system using verifying traces, e.g. \Shake, is unlikely to see
significant benefit from storing more than one \hs{Trace} per key\footnote{There
is a small chance of a benefit if the dependencies change but the result does
not, and then the dependencies change back to what they were before.}.
Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash v))},
where the initial \hs{k} is the \hs{key} field of \hs{Trace}.

>>>>

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it.
It also marks dirty all cells that (transitively) depend on the modified cell.
Excel does not record dynamic dependencies of each cell; instead it computes a
\emph{static over-approximation} -- it is safe for it to make more cells dirty
than necessary, but not vice versa. The over-approximation is as follows: a cell
is marked dirty if its formula statically refers to a dirty cell, or if the
formula calls a function like \cmd{INDIRECT} whose dependencies cannot be
guessed from the formula alone. The over-approximation is clear for
\cmd{INDIRECT}, but it is also present for \cmd{IF}, where both branches are
followed even though dynamically only one is used.

@@

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well if we skip any
number of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$. When $n \ge 2$, deep constructive traces require the tasks to be
\emph{deterministic}, as otherwise it is possible to violate correctness, as
illustrated by an example in~\S\ref{sec-cloud-aspects}.

% Concretely, if
% the result of building a key given a set of immediate dependencies can vary,
% then it is possible that key \hs{k} will have been computed locally, but
% something \emph{using} \hs{k} will have been taken as a constructive input using
% a different value for \hs{k}. If another rule combines those two values then two
% different values of \hs{k} will have been used in one rule, leading to problems
% such as link-errors for typical compilers -- a phenomenon known as a
% \emph{Frankenbuild}~\cite{esfahani2016cloudbuild}.

A downside of deep constructive traces is that they cannot support early cutoff
(\S\ref{sec-background-shake}), other than at $n$ levels of dependencies. On the
other hand, these traces are particularly useful for \emph{shallow builds}, as
discussed in~\S\ref{sec-cloud-aspects}.

@@

Here \hs{r} is either \hs{Hash}~\hs{v} (verifying traces) or
\hs{v} (constructive traces). A real system is highly likely to use a more
optimised implementation. Some of the most obvious optimisations are:

\begin{enumerate}
\item Any system using verifying traces is unlikely to see significant benefit
from storing more than one \hs{Trace} per key\footnote{There is a small chance
of a benefit if the dependencies change but the result does not, and then the
dependencies change back to what they were before.}. Therefore, such systems can
store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
initial \hs{k} is the \hs{key} field of \hs{Trace}.

### 2018-07-06 paper/5-implementations.tex
Minor revision

The \hs{modTimeRebuilder} uses the pair
\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
build information, carried by a \hs{MonadState} monad. This
\hs{MakeInfo} comprises the \emph{current time} \hs{now}~\hs{::}~\hs{Time} and the map
\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
modification times}.
We assume that the external system, which invokes the build system, updates
\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.

>>>>

The \hs{modTimeRebuilder} uses the pair
\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
build information, carried by a state monad. This \hs{MakeInfo} comprises
the \emph{current time} \hs{now}~\hs{::}~\hs{Time} and the map
\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file modification
times}. We assume that the external system, which invokes the build system,
updates \hs{MakeInfo} reflecting any file changes between successive builds.
